Effectful Semantics for Substructural SSA

### 2023-11-01 pldi24/paper.tex
Tap

@@

In a standard SSA-based compiler, to represent a function, we begin by decomposing it into basic blocks of the following form:
\TODO{this}
A program written in this manner being in ``SSA form'' simply means that every variable name is defined exactly once. For example,
\TODO{this}
is not in SSA form, but can easily be converted into SSA form by rewriting it as
\TODO{this}
Unfortunately, things get a little more tricky when we try to define variables in loops or branches. For example, consider the following code:
\TODO{this}
There are two competing definitions of x in the left and right blocks, and it’s unclear how to disambiguate between them. The usual solution is to introduce a \(\phi\)-node. In particular, we can write:
\TODO{this}
\TODO{segue}
The semantics of SSA, however, can be quite complex to reason about, especially when considering the somewhat unintuitive scoping rules of variables and \(\phi\)-function arguments.

>>>>

\usepackage{listings}
\usepackage{xcolor}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
%    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2
}

\lstset{style=mystyle}

@@

\TODO{text...}

\section{Background}

In a standard SSA-based compiler, to represent a function, we begin by decomposing it into basic blocks of the following form:
\begin{lstlisting}
'label:
  var = instruction
  ...
  var = instruction
  terminator
\end{lstlisting}
A program written in this manner being in ``SSA form'' simply means that every variable name is defined exactly once. For example,
\begin{lstlisting}
'entry:
  x = 7
  x = add x 2
  ret x
\end{lstlisting}
is not in SSA form, but can easily be converted into SSA form by rewriting it as
\begin{lstlisting}
'entry:
  x0 = 7
  x1 = add x0 2
  ret x1
\end{lstlisting}
Unfortunately, things get a little more tricky when we try to define variables in loops or branches. For example, consider the following code:
\begin{lstlisting}
'entry:
  br arg0 'left `right
'left:
  x = 7
  jmp 'exit
'right:
  x = 3
  jmp 'exit
'exit:
  ret x
\end{lstlisting}
There are two competing definitions of \(x\) in the left and right blocks, and it’s unclear how to disambiguate between them. The usual solution is to introduce a \(\phi\)-node. In particular, we can write:
\begin{lstlisting}
'entry:
  br arg0 'left `right
'left:
  x0 = 7
  jmp 'exit
'right:
  x1 = 3
  jmp 'exit
'exit:
  x2 = phi 'left x0 'right x1
  ret x2
\end{lstlisting}
\TODO{make above into graphs?}
\TODO{segue}
The semantics of SSA, however, can be quite complex to reason about, especially when considering the somewhat unintuitive scoping rules of variables and \(\phi\)-function arguments.
